---
title: "Crimes and Clearances"
author: "Nandita"
date: "2024-11-22"
output: pdf_document
---
Assembly Bill 109 was passed in 2011, shifting the responsibility of housing non-violent and
non-serious offenses from state facilities to county jails. Utilizing a difference-in-difference model where
Los Angeles, Orange, and Riverside Counties are treated and other units are control we perform a
difference-in-difference analysis where our outcome variable is Property crime

```{r setup, include=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(plm)
library(stargazer)
library(ggplot2)
library(tinytex)

# load in the dataset
crime_data <- read_csv("Crimes_and_Clearances_with_Arson-1985-2023 (1).csv")
```

The raw data represents county-level crime statistics aggregated annually. 

We aggregate the data to the county-by-year level summing up the data grouping county and year.
```{r, warning=FALSE}
crime_data <- crime_data %>%
  group_by(County, Year) %>%
  summarise(across(ends_with("sum"), sum, na.rm = TRUE), .groups = 'drop')
```
Since the variables already represent sums at the lower level, it is appropriate
to find the total count of crimes at the county level. Finding the average will 
dilute the actual crime counts.


First, we generate a plot to identify if the parallel trends assumption is satisfied.

```{r, warning=FALSE}
# assign values 1, 0 to treated and control groups
crime_data$treated_county <- ifelse(crime_data$County %in% 
                                      c("Los Angeles County", "Orange County",
                                        "Riverside County"), 1, 0)
crime_data$treated_time <- ifelse(crime_data$Year >= 2011, 1, 0)

# Interaction term for the difference-in-difference model
crime_data$treated <- crime_data$treated_county * crime_data$treated_time

# Create a factor variable for treated and control groups
crime_data$treated_county <- factor(crime_data$treated_county, levels = c(0, 1),
                                    labels = c("Control", "Treated"))

# calculate average property crime rate across counties to compare trends 
#between the treated group and control
property_data <- crime_data %>%
  group_by(Year, treated_county) %>%
  mutate(avg_crime_rate = sum(Property_sum)/n ())

# Plot to check parallel trends
ggplot(data = property_data, aes(x = Year, y = avg_crime_rate, color = 
                                   treated_county, group = treated_county)) +
  geom_line(size = 1.2) +  
  geom_vline(xintercept = 2011, linetype = "dashed", color = "gray40", 
             size = 0.8) +
  scale_color_manual(values = c("Control" = "#0072B2", "Treated" = "#D55E00")) + 
  labs(
    title = "Difference-in-Differences - Parallel Trends Assumption",
    x = "Year",
    y = "Averagae Property Crime Rate",
    color = "Group"
  ) + 
  theme_minimal(base_size = 14) + 
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),  
    legend.position = "top",  
    legend.title = element_text(face = "bold") 
  )
```
The plot suggests that the parallel trends assumption is violated, meaning that
the treated counties were not following a similar trend to the control counties 
before the passing of Assembly Bill 109.

This violation weakens the validity of a difference-in-differences analysis,
as the assumption that the treated and control counties would have followed the 
same trend in the absence of the policy change is no longer valid. 

```{r, warning=FALSE,results='asis'}
model <- plm(avg_crime_rate ~ treated_county*treated_time,index = c("County", 
                                                                    "Year")
             ,data=property_data )

stargazer(model, type = "latex", title = "Difference-in-Differences Model with 
          Two-Way Fixed Effects",
          dep.var.labels = "Average Crime Rate",
          covariate.labels = c("Treated Time",
                               "Treated Location x Treated Time"),
          omit.stat = c("f", "ser"), 
          no.space = TRUE)
```
On average, property crime rates across all counties decreased by approximately
4,076 crimes after 2011, regardless of whether they were in treated counties or 
not. The result is statistically significant (p < 0.01). After the policy, 
treated counties experienced an average additional reduction of 54,783 property 
crimes, compared to control counties. This result is also statistically 
significant (p < 0.01). From the r-squared value we can interpret that about 
37.4 percent of the variation in the model is explained by average property 
crime. However, since the parallel trends assumption is violated, 
we assume that the results are suggestive but not conclusive.
```

